---
title: "Comparing Earthworm Biomass and Invasive Species Density Across Soil Types"
subtitle:
author: "Rachel Krick"
institute: "University of Illinois at Chicago"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(broom)
library(lubridate)
library(janitor)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
invasivesp <- read_csv("../data/Mueller_buckthorn_gradient_JAE_2018_Dryad_data.csv")%>%
  clean_names()%>%
  drop_na()
```

```{r slide-style,  include=FALSE}
style_xaringan(
  title_slide_background_image = "img/Rhamnus-cathartica-fruits.jpg",
  title_slide_text_color = "white"
)
```


# Introduction & My Data

**Study looked at how invasive plant species influence ecological properties**

Two different methods of sampling:  

  * (1) **large-scale stratified sampling** of vegetation  across multiple sites where researchers recorded species composition, abundance, and environmental variables in areas with and without invasive plants
  
  * (2) **finer-scale paired-plot studies**, where invaded plots were directly matched with nearby uninvaded plots under similar conditions to better isolate invasion effects. 

##*Rhamnus cathartica*
- a deciduous invasive shrub with a non-native range that includes most of the United States and Canada 
- commonly known as **Buckthorn**
- unique because its leaves have higher concentrations of nutrients, develop earlier in spring, and senesce later in fall
- studies typically observe greater nutrient pools and fluxes in invaded stands 

---
#My Variables
- **`buckthorn_stems_per_m2_subplot`**: the number of buckthorn stems in each 3 m diameter subplot
 > measures the density of invasive buckthorn plants

- **`earthworm_biomass_g_per_m2`**: biomass of earthworms in each subplot

- **`soil_type`**: plot soil types, either sandy or silty


---
class: center, middle

# **Research Questions:**  

  1. Is there a statistically significant relationship between the invasive species and the earth worm biomass? What does this mean for how the invasive species impacts earthworm biomass? 
  
  2. Does the mean earthworm biomass vary by soil type? Is there a statistically significant difference in the means? 
  
  3. Does the mean number of buckthorn stems vary by soil type? Is there a statistically significant difference in the means? 

## **Overall:**  
 What does this mean for how soil type might play a role in the relationship between invasive species and earthworm biomass? 


---

class: center, middle
#Visualizing the Relationship Between Buckthorn and Earthworm Biomass

.pull-left[
```{r smooth_lm, echo=FALSE, fig.width= 6, out.width="90%"}

ggplot(invasivesp, aes(x =buckthorn_stems_per_m2_subplot, y = earthworm_biomass_g_per_m2)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color="darkgreen") +
  labs(
    x = "Buckthorn Stems per m¬≤",
    y = "Earthworm Biomass (g/m¬≤)",
    title = "Relationship between Earthworm Biomass & Buckthorn Stem Density"
  ) +
  theme_classic(base_size = 9)
```
]


.pull-right[
```{r exploratory_scatter, echo=FALSE, eval = TRUE, fig.width= 6, out.width="90%"}

ggplot(invasivesp, aes(x =buckthorn_stems_per_m2_subplot, y = earthworm_biomass_g_per_m2 , color = soil_type)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Buckthorn Stems per m¬≤",
    y = "Earthworm Biomass (g/m¬≤)",
    color = "soil_type",
    title = "Relationship between Earthworm Biomass & Buckthorn Stem Density by Soil Type"
  ) +
  theme_classic(base_size = 9)
```
]

---

class: center, middle
# Visualization of Difference in Means

.pull-left[
```{r two_sample_plot, echo=FALSE, eval=TRUE,  fig.width= 6, out.width="90%"}
ggplot(invasivesp, aes(x = soil_type, y = earthworm_biomass_g_per_m2)) +
  geom_boxplot() +
  labs(
    title = "Earthworm Biomass by Soil Type",
    x = "Soil Type",
    y = "Earthworm Biomass (g/m¬≤)"
  ) +
  theme_classic(base_size = 14)
```
]

.pull-right[
```{r two_sample_plot2, echo=FALSE, eval=TRUE,  fig.width= 6, out.width="90%"}
ggplot(invasivesp, aes(x = soil_type, y = buckthorn_stems_per_m2_subplot)) +
  geom_boxplot() +
  labs(
    title = "Buckthorn Stem Density by Soil Type",
    x = "Soil Type",
    y = "Buckthorn Stems per m¬≤"
  ) +
  theme_classic(base_size = 14)
```
]
---
**T-Test:**
Does the mean earthworm biomass vary by soil type? Is there a statistically significant difference in the means? 

.pull-left[

```{r choose_two_sample, echo=TRUE, eval=FALSE}
x1 <- invasivesp%>%
  filter(soil_type == "silty") %>%
  pull(earthworm_biomass_g_per_m2)

x2 <- invasivesp%>%
  filter(soil_type == "sandy") %>%
  pull(earthworm_biomass_g_per_m2)

```

```{r two_sample_summary, echo=TRUE, eval=FALSE}
xbar1 <- mean(x1, na.rm = TRUE)
xbar2 <- mean(x2, na.rm = TRUE)
s1 <- sd(x1, na.rm = TRUE)
s2 <- sd(x2, na.rm = TRUE)
n1 <- sum(!is.na(x1))
n2 <- sum(!is.na(x2))
mean_diff <- xbar1 - xbar2
tibble(
  xbar1 = xbar1,
  xbar2 = xbar2,
  mean_diff = mean_diff
)
```
]

.pull-right[
```{r two_sample_ttest, echo=TRUE, eval=FALSE}
t.test(x1, x2)
```

- xŐĄ1=8.32	
- xŐĄ2=6.814286	
- xŐĄ‚āĀ ‚ąí xŐĄ‚āā= 1.505714	
- t= 1.096
- df= 9.2415
- p-value= 0.3005
- CI= -1.587664 to 4.599092


]

---

**T-Test:** 
Does the mean number of buckthorn stems vary by soil type? Is there a statistically significant difference in the means? 

.pull-left[
```{r choose_two_sample2, echo=TRUE, eval=FALSE}
y1 <- invasivesp %>%
  filter(soil_type == "silty") %>%
  pull(buckthorn_stems_per_m2_subplot)

y2 <- invasivesp %>%
  filter(soil_type == "sandy") %>%
  pull(buckthorn_stems_per_m2_subplot)

```

```{r two_sample_summary2, echo=TRUE, eval=FALSE}
ybar1 <- mean(y1, na.rm = TRUE)
ybar2 <- mean(y2, na.rm = TRUE)
s1 <- sd(y1, na.rm = TRUE)
s2 <- sd(y2, na.rm = TRUE)
n1 <- sum(!is.na(y1))
n2 <- sum(!is.na(y2))
mean_diff <- ybar1 - ybar2
tibble(
  ybar1 = ybar1,
  ybar2 = ybar2,
  mean_diff = mean_diff
)
```
]

.pull-right[
```{r two_sample_ttest2, echo=TRUE, eval=FALSE}
t.test(y1, y2)
```

- »≥1= 9.94	
- »≥2= 3.914286	
- »≥‚āĀ ‚ąí »≥‚āā = 6.025714
- t=  2.1897
- df= 4.5444
- p-value= 0.08557
- CI= -1.266736 to 13.318164

]

---

# Results/Analysis
##Earth Worm Biomass
- The mean earthworm biomass was higher in silty soil than in sandy soil.
- Because 0 is included in the interval, there is not strong evidence of a significant difference between the   two soil types.
- The test result was not statistically significant.
> Because the p-value is greater than 0.05, we fail to reject the null hypothesis.

- This indicates there is insufficient evidence to conclude that earthworm 
  biomass differs significantly between silty and sandy soils.
  
---
# Results/Analysis
## Buckthorn Stem Density 
- The mean number of buckthorn stems was higher in silty soil than in sandy soil.
- Because 0 is included in the interval, there is not strong evidence of a significant difference between the two soil types.
- The test result was also not statistically significant.
> Because the p-value is greater than 0.05, we fail to reject the null hypothesis.

- This indicates there is insufficient evidence to conclude that buckthorn stem density differs significantly between the two soil types.

---

# Conclusions
- Soil type alone may not be a strong predictor of the relationship between invasive buckthorn and earthworm biomass in this study. 

- Both variables were higher in silty soils: this may indicate that silty soils provide more favorable conditions for both earthworms and buckthorn growth. 
  - Higher mean values in silty soils may reflect more favorable moisture and nutrient conditions